Fractional Types

### 2019-11-20 fracGC/p.tex
clean up code; move tikz to main text

\usepackage{booktabs}  
\usepackage{subcaption} 
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{bbm}
\usepackage[greek,english]{babel}
\usepackage{ucs}
\usepackage[utf8x]{inputenc}
\usepackage{autofe}
\usepackage[references]{agda} 
\usepackage{newunicodechar}
\usepackage[nocenter]{qtree}
\usepackage{tikz}
\usepackage{bbold}
\usetikzlibrary{cd}
\usetikzlibrary{quotes}

\newunicodechar{𝕌}{$\mathbb{U}$}
\newunicodechar{𝔹}{$\mathbb{B}$}
\newunicodechar{𝔽}{$\mathbb{F}$}
\newunicodechar{𝕋}{$\mathbb{T}$}
\newunicodechar{𝟘}{$\mathbb{0}$}
\newunicodechar{𝟙}{$\mathbb{1}$}
\newunicodechar{𝟚}{$\mathbb{2}$}
\newunicodechar{⋆}{${}_*$}
\newunicodechar{○}{$\gcv$}
\newunicodechar{ᵤ}{${}_u$}
\newunicodechar{⊚}{$\fatsemi$}
\newunicodechar{⟷}{$\leftrightarrow$}
\newunicodechar{⟶}{$\longrightarrow$}
\newunicodechar{⟪}{$\ll$}
\newunicodechar{⟫}{$\gg$}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[t]
\[\begin{array}{rrcll}
\idc :& \tau & \iso & \tau &: \idc \\
\\
\identlp :&  0 + \tau & \iso & \tau &: \identrp \\
\swapp :&  \tau_1 + \tau_2 & \iso & \tau_2 + \tau_1 &: \swapp \\
\assoclp :&  \tau_1 + (\tau_2 + \tau_3) & \iso & (\tau_1 + \tau_2) + \tau_3 &: \assocrp \\
\\
\identlt :&  1 * \tau & \iso & \tau &: \identrt \\
\swapt :&  \tau_1 * \tau_2 & \iso & \tau_2 * \tau_1 &: \swapt \\
\assoclt :&  \tau_1 * (\tau_2 * \tau_3) & \iso & (\tau_1 * \tau_2) * \tau_3 &: \assocrt \\
\\
\distz :&~ 0 * \tau & \iso & 0 ~ &: \factorzl \\
\dist :&~ (\tau_1 + \tau_2) * \tau_3 & \iso & (\tau_1 * \tau_3) + (\tau_2 * \tau_3)~ &: \factor
\end{array}\]
\begin{center}
\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_2 \iso \tau_3}
{\jdg{}{}{c_1 \odot c_2 : \tau_1 \iso \tau_3}}
{}
\\
\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_3 \iso \tau_4}
{\jdg{}{}{c_1 \oplus c_2 : \tau_1 + \tau_3 \iso \tau_2 + \tau_4}}
{}
\qquad
\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_3 \iso \tau_4}
{\jdg{}{}{c_1 \otimes c_2 : \tau_1 * \tau_3 \iso \tau_2 * \tau_4}}
{}
\end{center}
\caption{$\Pi$-terms and combinators.}
\label{pi-terms}
\end{figure*}

@@

\noindent The top path is the $\Pi$ program
$(c_1~\oplus~c_2)~\odot~\swapp$ which acts on the type $A$ by $c_1$,
acts on the type $B$ by $c_2$, and deforms the resulting space by a
twist that exchanges the two injections into the sum type. The bottom
path performs the twist first and then acts on the type $A$ by $c_1$
and on the type $B$ by $c_2$ as before. If one could imagine the paths
as physical wires and the actions $c_1$ and $c_2$ as permutations on
these wires then, holding the points $A$, $B$, $C$, and $D$ fixed, it
is possible to rotate the top part of the diagram to become identical
to the bottom one. That rotation can be undone (reversed), which takes
the bottom part of the diagram into the top part.  In other words,
there exists a continuous deformation of the program
$(c_1~\oplus~c_2)~\odot~\swapp$ to the program
$\swapp \odot (c_2~\oplus~c_1)$. We can also show that this means
that, as permutations, $(c_1~\oplus~c_2)~\odot~\swapp$ and
$\swapp \odot (c_2~\oplus~c_1)$ are equal. This relation is
non-trivial, as not all programs between the same types can be
deformed into one another. The simplest example of inequivalent
deformations are the two automorphisms of $1+1$, namely $\idc$ and
$\swapp$.

@@

A conventional and efficient way to ensure that every allocation is
paired with a matching de-allocation is to impose a stack
discipline. An illustrative demonstration of this idea is the
\verb|with_ancilla| operator from
Quipper~\cite{Green:2013:QSQ:2491956.2462177}:
\begin{verbatim}
with_ancilla :: (Qubit -> Circ a) -> Circ a
\end{verbatim}
The operator takes a block of gates parameterized by an ancilla,
allocates a new ancilla of type \verb|Qubit| initialized to $\ket{0}$,
and runs the given block of gates. At the end of its execution, the
block is expected to return the ancilla to the state $\ket{0}$ (this
is the second subproblem addressed later) at which point it is
de-allocated. In this scoped-by-construction approach, allocation and
de-allocation of ancillae requires nothing beyond conventional
parameter-passing techniques in which the parameter is allocated
before entry to a function and de-allocated at exit.

@@

To make this idea work, several technical details are needed. Most
notably, we must exclude the empty types from this creation and
annihilation process. Otherwise, we would be able to prove that:
\[\begin{array}{rcll}
1 &=& 0 * 1/0 & \textrm{by~} \eta_0 \\
&=& 0 & \textrm{by~} \absorbr
\end{array}\]
We therefore extend the syntax of core $\Pi$ in Sec.~\ref{sub:core} as follows:
\[\begin{array}{lrcl}
\textit{Value types} & t &::=& \cdots \alt \oneover{t} \\
\textit{Non-empty types} & \net \\
\textit{Values}      & v &::=& \cdots \alt \gcv \\
\textit{Level-1 types} &&& t \leftrightarrow t \\
\textit{Level-1 programs} & c &::=& \cdots \alt
   \eta : 1 \leftrightarrow (\net * \oneover{\net}) \\
   &&& ~~~~~\alt \epsilon : (\net * \oneover{\net}) \leftrightarrow 1 \\
\textit{Level-2 types} &&& c \Leftrightarrow c \\
\textit{Level-2 programs} & \alpha &::=& \cdots 
  \alt \mathit{inv} : \epsilon \circ \eta \Leftrightarrow \idc 
\end{array}\]

@@

@@

% of course need de-allocation; keep track of lifetime

% easy: stack but not general

% for general; programmer control or gc

@@

@@

@@

We use a Singleton type
to remember this information.

Formalize ancilla problem in the context of Pi using eta and epsilon
as done in PiFracDyn; more general than scoped; still runtime
check.

>>>>

\usepackage{booktabs}  
\usepackage{subcaption} 
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{bbm}
\usepackage[greek,english]{babel}
\usepackage{ucs}
\usepackage[utf8x]{inputenc}
\usepackage{autofe}
\usepackage[references]{agda} 
\usepackage{newunicodechar}
\usepackage[nocenter]{qtree}
\usepackage{multicol}  
\usepackage{tikz}
\usepackage{bbold}
\usetikzlibrary{cd}
\usetikzlibrary{quotes}

\newunicodechar{≟}{$\mathbin{\stackrel{?}{=}}$}
\newunicodechar{𝕌}{$\mathbb{U}$}
\newunicodechar{𝔹}{$\mathbb{B}$}
\newunicodechar{𝔽}{$\mathbb{F}$}
\newunicodechar{𝕋}{$\mathbb{T}$}
\newunicodechar{𝟘}{$\mathbb{0}$}
\newunicodechar{𝟙}{$\mathbb{1}$}
\newunicodechar{𝟚}{$\mathbb{2}$}
\newunicodechar{⋆}{${}_*$}
\newunicodechar{○}{$\gcv$}
\newunicodechar{ᵤ}{${}_u$}
\newunicodechar{⊚}{$\fatsemi$}
\newunicodechar{⟷}{$\leftrightarrow$}
\newunicodechar{⟶}{$\longrightarrow$}
\newunicodechar{⟪}{$\ll$}
\newunicodechar{⟫}{$\gg$}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[t]
\begin{multicols}{2}
\[\begin{array}{rrcll}
\idc :& \tau & \iso & \tau &: \idc \\
\\
\identlp :&  0 + \tau & \iso & \tau &: \identrp \\
\swapp :&  \tau_1 + \tau_2 & \iso & \tau_2 + \tau_1 &: \swapp \\
\assoclp :&  \tau_1 + (\tau_2 + \tau_3) & \iso & (\tau_1 + \tau_2) + \tau_3 &: \assocrp \\
\\
\identlt :&  1 * \tau & \iso & \tau &: \identrt \\
\swapt :&  \tau_1 * \tau_2 & \iso & \tau_2 * \tau_1 &: \swapt \\
\assoclt :&  \tau_1 * (\tau_2 * \tau_3) & \iso & (\tau_1 * \tau_2) * \tau_3 &: \assocrt \\
\\
\distz :&~ 0 * \tau & \iso & 0 ~ &: \factorzl \\
\dist :&~ (\tau_1 + \tau_2) * \tau_3 & \iso & (\tau_1 * \tau_3) + (\tau_2 * \tau_3)~ &: \factor
\end{array}\]
\vspace{2cm}
\begin{center}
\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_2 \iso \tau_3}
{\jdg{}{}{c_1 \fatsemi c_2 : \tau_1 \iso \tau_3}}
{}

\vspace{1cm}

\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_3 \iso \tau_4}
{\jdg{}{}{c_1 \oplus c_2 : \tau_1 + \tau_3 \iso \tau_2 + \tau_4}}
{}

\vspace{1cm}

\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_3 \iso \tau_4}
{\jdg{}{}{c_1 \otimes c_2 : \tau_1 * \tau_3 \iso \tau_2 * \tau_4}}
{}
\end{center}
\end{multicols}
\caption{$\Pi$-terms and combinators.}
\label{pi-terms}
\end{figure*}

@@

\noindent The top path is the $\Pi$ program
$(c_1~\oplus~c_2)~\fatsemi~\swapp$ which acts on the type $A$ by $c_1$,
acts on the type $B$ by $c_2$, and deforms the resulting space by a
twist that exchanges the two injections into the sum type. The bottom
path performs the twist first and then acts on the type $A$ by $c_1$
and on the type $B$ by $c_2$ as before. If one could imagine the paths
as physical wires and the actions $c_1$ and $c_2$ as permutations on
these wires then, holding the points $A$, $B$, $C$, and $D$ fixed, it
is possible to rotate the top part of the diagram to become identical
to the bottom one. That rotation can be undone (reversed), which takes
the bottom part of the diagram into the top part.  In other words,
there exists a continuous deformation of the program
$(c_1~\oplus~c_2)~\fatsemi~\swapp$ to the program
$\swapp \fatsemi (c_2~\oplus~c_1)$. We can also show that this means
that, as permutations, $(c_1~\oplus~c_2)~\fatsemi~\swapp$ and
$\swapp \fatsemi (c_2~\oplus~c_1)$ are equal. This relation is
non-trivial, as not all programs between the same types can be
deformed into one another. The simplest example of inequivalent
deformations are the two automorphisms of $1+1$, namely $\idc$ and
$\swapp$.

@@

A conventional and efficient way to ensure that every allocation is
paired with a matching de-allocation is to impose a stack
discipline. An illustrative demonstration of this idea is the
\verb|with_ancilla| operator from
Quipper~\cite{Green:2013:QSQ:2491956.2462177} :
\begin{verbatim}
with_ancilla :: (Qubit -> Circ a) -> Circ a
\end{verbatim}
The operator takes a block of gates parameterized by an ancilla,
allocates a new ancilla of type \verb|Qubit| initialized to $\ket{0}$,
and runs the given block of gates. At the end of its execution, the
block is expected to return the ancilla to the state $\ket{0}$ (this
is the second subproblem addressed later) at which point it is
de-allocated. In this scoped-by-construction approach, allocation and
de-allocation of ancillae requires nothing beyond conventional
parameter-passing techniques in which the parameter is allocated
before entry to a function and de-allocated at exit.

@@

To make this idea work, several technical details are needed. Most
notably, we must exclude the empty types from this creation and
annihilation process. Otherwise, we would be able to prove that:
\[\begin{array}{rcll}
1 &=& 0 * 1/0 & \textrm{by~} \eta_0 \\
&=& 0 & \textrm{by~} \absorbr
\end{array}\]
We therefore extend the syntax of core $\Pi$ in Sec.~\ref{sub:core} as follows:
\[\begin{array}{lrcl}
\textit{Value types} & t &::=& \cdots \alt \oneover{t} \\
\textit{Non-empty types} & \net \\
\textit{Values}      & v &::=& \cdots \alt \gcv \\
\textit{Level-1 types} &&& t \leftrightarrow t \\
\textit{Level-1 programs} & c &::=& \cdots \alt
   \eta : 1 \leftrightarrow (\net * \oneover{\net}) \\
   &&& ~~~~~\alt \epsilon : (\net * \oneover{\net}) \leftrightarrow 1 \\
\textit{Level-2 types} &&& c \Leftrightarrow c \\
\textit{Level-2 programs} & \alpha &::=& \cdots 
  \alt \mathit{inv} : \eta \fatsemi \epsilon \Leftrightarrow \idc 
\end{array}\]

@@

\tikzset{every picture/.style={line width=0.75pt}} %set default line width to 0.75pt        
\begin{tikzpicture}[x=0.75pt,y=0.75pt,scale=0.8, yscale=-1]
%uncomment if require: \path (0,300); %set diagram left start at 0, and has height of 300
%Straight Lines [id:da05403550183378514] 
\draw    (10,20) -- (190,20) ;
%Shape: Circle [id:dp8701819096818126] 
\draw   (138.75,21.25) .. controls (138.75,15.04) and (143.79,10) .. (150,10) .. controls (156.21,10) and (161.25,15.04) .. (161.25,21.25) .. controls (161.25,27.46) and (156.21,32.5) .. (150,32.5) .. controls (143.79,32.5) and (138.75,27.46) .. (138.75,21.25) -- cycle ;
%Straight Lines [id:da3470499764023487] 
\draw    (150,80) -- (150,10) ;
%Straight Lines [id:da9303973632982276] 
\draw    (40,80) -- (190,80) ;
%Shape: Circle [id:dp03942567054970991] 
\draw   (48.75,81.33) .. controls (48.75,75.12) and (53.79,70.08) .. (60,70.08) .. controls (66.21,70.08) and (71.25,75.12) .. (71.25,81.33) .. controls (71.25,87.55) and (66.21,92.58) .. (60,92.58) .. controls (53.79,92.58) and (48.75,87.55) .. (48.75,81.33) -- cycle ;
%Straight Lines [id:da08295302434157614] 
\draw    (60,92.58) -- (60,20) ;
%Shape: Arc [id:dp2904677718814479] 
\draw  [draw opacity=0] (38.49,110.13) .. controls (28.3,109.22) and (20.27,102.7) .. (20.15,94.89) .. controls (20.02,86.57) and (28.89,79.93) .. (40,80) -- (40.38,95.12) -- cycle ; \draw   (38.49,110.13) .. controls (28.3,109.22) and (20.27,102.7) .. (20.15,94.89) .. controls (20.02,86.57) and (28.89,79.93) .. (40,80) ;
%Straight Lines [id:da32172184710713914] 
\draw    (38.49,110.13) -- (250,110) ;
%Shape: Arc [id:dp8398868072937593] 
\draw  [draw opacity=0] (248.77,109.75) .. controls (262.77,109.82) and (274.32,103.33) .. (274.64,95.17) .. controls (274.95,87.02) and (263.95,80.28) .. (250,80) -- (249.05,94.87) -- cycle ; \draw   (248.77,109.75) .. controls (262.77,109.82) and (274.32,103.33) .. (274.64,95.17) .. controls (274.95,87.02) and (263.95,80.28) .. (250,80) ;
%Straight Lines [id:da5220296307776525] 
\draw    (190,20) -- (250,80) ;
%Straight Lines [id:da320377013704656] 
\draw    (190,80) -- (250,20) ;
%Straight Lines [id:da8933694171914682] 
\draw    (250,20) -- (308,20) ;
\end{tikzpicture}

\begin{tikzpicture}[x=0.75pt,y=0.75pt,scale=0.8, yscale=-1]
%uncomment if require: \path (0,300); %set diagram left start at 0, and has height of 300
%Straight Lines [id:da9303973632982276] 
\draw    (40,80) -- (190,80) ;
%Shape: Arc [id:dp2904677718814479] 
\draw  [draw opacity=0] (38.49,110.13) .. controls (28.3,109.22) and (20.27,102.7) .. (20.15,94.89) .. controls (20.02,86.57) and (28.89,79.93) .. (40,80) -- (40.38,95.12) -- cycle ; \draw   (38.49,110.13) .. controls (28.3,109.22) and (20.27,102.7) .. (20.15,94.89) .. controls (20.02,86.57) and (28.89,79.93) .. (40,80) ;
%Straight Lines [id:da32172184710713914] 
\draw    (38.49,110.13) -- (250,110) ;
%Shape: Arc [id:dp8398868072937593] 
\draw  [draw opacity=0] (248.77,109.75) .. controls (262.77,109.82) and (274.32,103.33) .. (274.64,95.17) .. controls (274.95,87.02) and (263.95,80.28) .. (250,80) -- (249.05,94.87) -- cycle ; \draw   (248.77,109.75) .. controls (262.77,109.82) and (274.32,103.33) .. (274.64,95.17) .. controls (274.95,87.02) and (263.95,80.28) .. (250,80) ;
%Straight Lines [id:da7474565765110813] 
\draw    (40.09,30.3) -- (190.09,30.3) ;
%Shape: Arc [id:dp593084478238281] 
\draw  [draw opacity=0] (38.58,60.43) .. controls (28.39,59.52) and (20.36,53) .. (20.23,45.19) .. controls (20.11,36.87) and (28.98,30.23) .. (40.09,30.3) -- (40.47,45.42) -- cycle ; \draw   (38.58,60.43) .. controls (28.39,59.52) and (20.36,53) .. (20.23,45.19) .. controls (20.11,36.87) and (28.98,30.23) .. (40.09,30.3) ;
%Straight Lines [id:da5682447818267274] 
\draw    (38.58,60.43) -- (250.09,60.3) ;
%Shape: Arc [id:dp8936250935910023] 
\draw  [draw opacity=0] (248.86,60.05) .. controls (262.86,60.12) and (274.41,53.63) .. (274.72,45.47) .. controls (275.03,37.32) and (264.04,30.58) .. (250.09,30.3) -- (249.13,45.17) -- cycle ; \draw   (248.86,60.05) .. controls (262.86,60.12) and (274.41,53.63) .. (274.72,45.47) .. controls (275.03,37.32) and (264.04,30.58) .. (250.09,30.3) ;
%Straight Lines [id:da9423433355546923] 
\draw    (190,30) -- (250,80) ;
%Straight Lines [id:da8064437888434692] 
\draw    (190,80) -- (250.09,30.3) ;
\end{tikzpicture}

@@

% of course need de-allocation; keep track of lifetime
% easy: stack but not general
% for general; programmer control or gc

@@

%%%%% 
\subsection{State of the Art}

@@

Ricercar quote:
\begin{quote}
  Checking that some defined ancillae are indeed unchanged for all
  possible input vectors of an arbitrary description is hard; in
  general, one has to test all possible input vectors, which is
  undesirable. For a reversible programming language such as Janus
  [22], this is therefore imple- mented as a runtime check that checks
  the reversibility of a program only in rela- tion to the executing
  input vector. This is also the case for the syntax described in
  Fig. 1.

  To be able to introduce and remove ancilla wires from a circuit, we
  have identified the need for five basic rules.
\end{quote}

%%%%% 
\subsection{Choosing Ancillae Values}

@@
